Differential chemotaxis with competition as a mapping mechanism in the brain

### 2022-09-26 paper/main.tex
Monday WIP on paper discussion

% If no page numbering wanted:
%\pagenumbering{gobble}

@@

% Define title, empty date and authors
\title {
Time to stop; agent-based modelling of chemoaffinity with competition
}
\date{} % No date please
\author{\Authors}

@@

\emph{We show that a true chemoaffinity model based on gradient-following (via receptor-ligand interactions) combined with receptor-ligand-mediated axonal repulsion, in which information is only transferred by nearest-neighbour interactions, can account for a broad range of experimental manipulations of the retinotectal system.}

@@

% Could use chemoaffinity as a coverall theory, comprising chemospecificity and chemotaxis.

Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.
%
The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'. 
The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones. 
%
This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target.  
An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of gene expression spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity (\textbf{find citation}).
In this system a two-dimensional map of the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum. 
The retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds? 
%
The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
Two classes of ephrins have been identified, labelled `A' and `B'.
Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction. 
Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

@@

Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}. 
These models have also helped establish how competition between axons for space in the target tissue can additionally account for many of the known effects of surgical manipulation on map development.

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}. 
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue. 
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples). 
% Not sure about this:
However as the number of terms involved in PDEs necessarily grows large in systems representing differential chemotaxis, these systems quickly become impossible to analyse, making this potential advantage of such modelling redundant.
Additionally, the numerical simulation of a PDE system can become intractable, because each axonal projection must be represented by several state variables, each of which require solution across the entire spatial domain (which may consist of many thousands of elements).
% in james_modelling we did 41 projections. I found that trying to build a retinotectal system with about 100 projections was already intractable.

% But know we can motivate agent based modelling because *simulation* of PDEs for differential chemotaxis is intractable.
An alternative simulation approach is to treat each of $n$ axons as an agent that interacts with its environment and with the other $n-1$ axons. 
The computation of such a system (see for example, \citet{simpson_simple_2011}) is dominated by the $O(n^2)$ axon-axon interactions and so a limit will be approached as $n\rightarrow \infty$. 
We found that agent based simulations of $O(10^3)$ interacting axons can be achieved on a standard desktop computer whereas an equivalent PDE system would fail with about 100 axon projections.

\emph{PDEs too complex (can't do maths analysis), so need agent based. Can represent more biological detail and ask how that affects the systems level.}

\emph{List the questions that emerge from the discussion above}

\emph{Can Gierer model explain ALL the manipulations that S and G did?}

\emph{Allows us to make a specific representation of Gierer's model in which the receptor-ligand mechanism is elaborated.}

\textbf{xxxxxxxxxxxxxxx}

Computational models have shown that in addition to chemotaxis, local competitive interactions between growing axons can account for a wide range of results from experiments in which the factors known to influence the growth of axons have been surgically or genetically manipulated \citep{prestige_role_1975,simpson_simple_2011,suetterlin_target-independent_2014}.

However, while such models have \emph{assumed} the effects of chemotaxis, the key assumption of the chemoaffinity model, that axons sample and evaluate positional information only \emph{locally}, has yet to be tested directly.
%
Consider, for example, an important model by Goodhill \& Simpson, which shows how [a certain set of assumptions about the form of the] competition between axons can account for patterns of innervation in the developing tectum that result from ablating/rotating sections of the retina \citep{simpson_simple_2011}. 
%   
The formulation of this model includes a term that at every time-step moves the simulated axons in the direction of a pre-specified target location. 
Thus the model implicitly assumes a global supervisor, capable of evaluating the current state of each axon, computing a direction vector to correct its trajectory, and communicating that information to each axon.
%   
This model was constructed primarily to establish the effects of competition, and to this end is elegantly formulated.
%
Nevertheless, the key assumptions of the chemoaffinity model, and specifically the proposal that axons find their way by evaluating only local information (chemotaxis), are yet to be investigated directly.

The aim of the current study is therefore to construct a model that represents the assumptions of the chemotaxis model explicitly and to investigate how this mechanism behaves when combined with competitive interactions.

One advantage of such a model is that it can be used to investigate the sensitivity and robustness of local interactions in establishing topographic maps to various sources of noise.

Indeed, our results show that the local mechanisms of axonal information-processing that are assumed by the chemoaffinity model provide a remarkably robust solution to the problem of topographic map development.

@@

In an agent-based model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \mathbf{M}$ with the movement vector $\mathbf{M}$ given by
\begin{equation} \label{e:mv2}
 \mathbf{M} = \mathbf{G} +  \mathbf{X} + \mathbf{B}.
\end{equation}
$\mathbf{G}$ is the movement due to the chemotaxis effect and $\mathbf{X}$ is the movement due to competitive interactions with other branches. 
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue \citep{holt_target_1998}.

@@

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid. 
($N$ ligands, $l_i$, are also expressed by RGCs, but these are not used by this chemotaxis model.) 
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B).
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (Fig.\,\ref{f:ex} summarises results from the literature). 
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ includes the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.

@@

 We modelled competition as a distance-based competitive movement vector, $\mathbf{X}$, which is a sum of interactions that are `switched on' via receptor-ligand signalling:
%
\begin{equation} \label{e:X}
\mathbf{X} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$) and $B_{b}$ is the set of branches that are within interaction distance of $b$. $m_{\!_X}$ is a scalar parameter, $nM$ is the total number of branches and $\hat{\mathbf{x}}_{kb}$ is the unit vector from $k$ to $b$.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between two branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$):
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$r_{\!_X}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$. 
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
$s$ is a positive signal threshold for interaction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1. 
$Q$ is set to 1 only if at least one repulsive signal exceeds $s$ and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$.

@@

An assumption made in these papers is that all EphA receptor subtypes contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side. 
The precise contribution of each subtype to the overall level is assumed not to be important. 
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments. 
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki). 
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500. 
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue. 
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that they are drawn from statistically identical distributions.
% While the convergence of the manipulated and wildtype lines was less abrupt than in the experimental result of \citet{brown_topographic_2000}, it was comparable with the result from \citet{simpson_simple_2011}.

@@

The previous approach assumed that all EphA receptor subtypes (EphA3, EphA4, EphA5, etc) transmit the same signal to induce indistinguishable function.
This assumption is difficult to support, given that the effect caused by knock-\emph{in} of EphA3 (a shift of nasal RGC axons towards the rostral tectum) is enhanced by the knock-\emph{down} of EphA4. 
We therefore changed our assumption to the following: Most EphA receptors share the same function (EphA3, EphA5 etc) but EphA4 is different. 
We redefined $r_{\!\scriptscriptstyle 0,b}$ as the expression of EphA receptors (excluding EphA4) on branch $b$ (Fig.\,\ref{f:clustermech}A), and introduced $r_{\!\scriptscriptstyle A4,b}$ as the expression of EphA4 receptors (Fig.\,\ref{f:clustermech}B). EphA4 receptors are expressed in a spatially inhomogeneous pattern~\citep{reber_relative_2004} for which we choose the value $r_{\!\scriptscriptstyle A4} = 3.5$, dropping the branch subscript.  We assume that cis-interactions between retinal ephrinA ligands and retinal EphA4 receptors lead to bound and un-bound EphA4 receptors \citep{hornberger_modulation_1999}. The number of cis-bound EphA4 receptors is assumed to be $r_{\!\scriptscriptstyle A4}^{\mathrm{cis}} = w_{\!\scriptscriptstyle A4} \times r_{\!\scriptscriptstyle A4} \times l_0$ where $w_{\!\scriptscriptstyle A4} = 0.153$ is a binding weight modelling the probability that an EphA4 receptor will bind to a nearby retinal ephrinA ligand. Remaining, free EphA4 receptors are $r_{\!\scriptscriptstyle A4}^{\mathrm{free}} = r_{\!\scriptscriptstyle A4} - r_{\!\scriptscriptstyle A4}^{\mathrm{cis}}$ (Fig.\,\ref{f:clustermech}C).

It is known that EphA receptors form clusters whose presence is necessary for ligand attachments to induce signal transmission and that cluster size affects signal size \citep{nikolov_ephephrin_2013}. It is also known that that some receptor sub-types may compete for ligand attachment \citep{fiore_regulation_2019}.
Additionally, EphA4 receptors make `side attachments' to clusters of EphA receptors, possibly regulating the cluster size \citep{nikolov_ephephrin_2013}. 
%
We focused on the idea that signal strength is related to receptor cluster size and defined $c_i$, the effective cluster size for receptor $i$. 
We set $c_i=1$ for $i=1$, $2$ or $3$ and investigated a simple relationship between EphA cluster size, $c_0$, and $r_{\!\scriptscriptstyle A4}$, the density of EphA4 receptors:
\begin{equation}
    c_0 = 1/r_{\!\scriptscriptstyle A4}
\end{equation}
%
Eq.\,\ref{e:G} was modified to include the cluster size:
%
\begin{equation}\label{e:Gcs}
\mathbf{G} = m_{\!_G}\,\sum_i^N F_i\,c_i\,r_{i,b} \nabla L_i(\mathbf{x})
\end{equation}

@@

>>>>

@@

\title {
Differential chemotaxis with competition as a mapping mechanism in the brain
}
\date{} % No date please
\author{\Authors}

@@

\emph{The retinotectal projection, in which the axons of retinal ganglion cells form an ordered map in the optic tectum, is an important developmental system that has been studied for many decades. There is a large body of evidence, consisting of surgical and genetic manipulations, that give insight into how it behaves. Experimental work has uncovered a number of mechanisms that are involved in aspects of the mapping. However, a single, consistent theoretical framework has yet to describe all the patterns that are formed. Sperry's chemoaffinity theory provides an account of the wildtype pattern. Gierer formalised and extended this theory to incorporate competition to account for a subset of surgical manipulation experiments, but it was never tested against the full range of surgical manipulations. Other theoretical and computational studies have been partially successful in accounting for surgical manipulations, but to date, no single theory, operating only by local interactions, has accounted for every manipulation. We present a two-dimensional model of the retinotectal projection resembling that of Gierer, in which orthogonal pairs of opposing tectal ligand gradients combined with retinal gradients of receptors form Gierer's potential functions and show that this receptor-ligand model accounts for the full range of surgical manipulations. We then extend the detail with which the receptor-ligand interactions are treated to account for more recent genetic manipulation experiments.  }

@@

Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.
%
The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'. 
The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones. 
%
This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target.  
An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of chemicals spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity (\textbf{find citation}).
In this system a two-dimensional map of the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum. 
This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds? 
%
The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo studies \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002}.
Two classes of ephrins have been identified, labelled `A' and `B'.
Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction. 
Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

@@

Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}. 
The model helped establish how \emph{competition between axons} for space in the target tissue can additionally account for  map development after tissue ablation.

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}. 
%\emph{PDEs too complex (can't do maths analysis), so need agent based. Can represent more biological detail and ask how that affects the systems level.}
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue. 
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples). 
However as the number of terms involved in PDEs necessarily grows large in systems representing differential chemotaxis, their analytical tractability quickly disappears, making the potential advantage of such modelling redundant.
Additionally, the numerical simulation of a PDE system can become intractable, because each axonal projection must be represented by several state variables, each of which require solution across the entire spatial domain (which may consist of many thousands of elements).

An alternative simulation approach is to treat each of $n$ axons as an agent that interacts with its environment and with the other $n-1$ axons. 
The computation of such a system (see for example, \citet{simpson_simple_2011}) is dominated by the $O(n^2)$ axon-axon interactions and so a limit will be approached as $n\rightarrow \infty$. 
We found that agent based simulations of $O(10^3)$ interacting axons can be achieved on a standard desktop computer whereas an equivalent PDE system would fail with about 100 axon projections.
A further advantage of an agent based simulation is that the addition of new biological details is easy to implement and comes an a minimal computational cost, whereas in a PDE system, each new detail requires at least one state variable, which must be solved across the entire domain.

% List questions that emerge from the above
For these reasons, we sought to implement an agent-based differential chemotaxis model of the retinotectal projection to ask: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

@@

In an agent-based model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \mathbf{M}$ with the movement vector $\mathbf{M}$ given by
\begin{equation} \label{e:M}
 \mathbf{M} = \mathbf{G} +  \mathbf{X} + \mathbf{B}.
\end{equation}
$\mathbf{G}$ is the movement due to the chemotaxis effect and $\mathbf{X}$ is the movement due to competitive interactions with other branches. 
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue \citep{holt_target_1998}.

@@

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid. 
($N$ ligands, $l_i$, may also be expressed by RGCs, but these are not used by the chemotaxis model.) 
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B). Thus the model is of \emph{differential chemotaxis}.
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (Supplementary Fig.\,\ref{f:experimental_expressions} summarises results from the literature). 
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ could include EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ could include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.

@@

 We modelled competition as a distance-based competitive movement vector, $\mathbf{X}$, which is a sum of interactions that are `switched on' via receptor-ligand signalling:
%
\begin{equation} \label{e:X}
\mathbf{X} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$r_{\!_X}$ is the receptor-ligand interaction radius, $W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$) and $B_{b}$ is the set of branches that are within interaction distance ($2 r_{\!_X}$) of $b$. $m_{\!_X}$ is a scalar parameter, $nM$ is the total number of branches and $\hat{\mathbf{x}}_{kb}$ is the unit vector from $k$ to $b$.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between two branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$):
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$. 
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
$s$ is a positive signal threshold for interaction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1. 
$Q$ is set to 1 only if at least one repulsive signal exceeds $s$ and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$.

@@

An assumption made in these papers is that all EphA receptor sub-types contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side. 
The precise contribution of each sub-type to the overall level is assumed not to be important. 
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments. 
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki). 
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500. 
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue. 
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that they are drawn from statistically identical distributions.
% While the convergence of the manipulated and wildtype lines was less abrupt than in the experimental result of \citet{brown_topographic_2000}, it was comparable with the result from \citet{simpson_simple_2011}.

@@

The previous approach assumed that all EphA receptor sub-types (EphA3, EphA4, EphA5, etc) transmit the same signal to induce indistinguishable function.
This assumption is difficult to support, given that the effect caused by knock-\emph{in} of EphA3 (a shift of nasal RGC axons towards the rostral tectum) is enhanced by the knock-\emph{down} of EphA4. 
We therefore changed our assumption to the following: Most EphA receptors share the same function (EphA3, EphA5 etc) but EphA4 is different. 
We redefined $r_{\!\scriptscriptstyle 0,b}$ as the expression of EphA receptors (excluding EphA4) on branch $b$ (Fig.\,\ref{f:clustermech}A), and introduced $r_{\!\scriptscriptstyle A4,b}$ as the expression of EphA4 receptors (Fig.\,\ref{f:clustermech}B). EphA4 receptors are expressed in a spatially inhomogeneous pattern~\citep{reber_relative_2004} for which we choose the value $r_{\!\scriptscriptstyle A4} = 3.5$, dropping the branch subscript.  We assume that cis-interactions between retinal ephrinA ligands and retinal EphA4 receptors lead to bound and un-bound EphA4 receptors \citep{hornberger_modulation_1999}. The number of cis-bound EphA4 receptors is assumed to be $r_{\!\scriptscriptstyle A4}^{\mathrm{cis}} = w_{\!\scriptscriptstyle A4} \times r_{\!\scriptscriptstyle A4} \times l_0$ where $w_{\!\scriptscriptstyle A4} = 0.153$ is a binding weight modelling the probability that an EphA4 receptor will bind to a nearby retinal ephrinA ligand. Remaining, free EphA4 receptors are $r_{\!\scriptscriptstyle A4}^{\mathrm{free}} = r_{\!\scriptscriptstyle A4} - r_{\!\scriptscriptstyle A4}^{\mathrm{cis}}$ (Fig.\,\ref{f:clustermech}C).

It is known that EphA receptors form clusters, that these are necessary for ligand attachments to induce signal transmission and that cluster size affects signal size \citep{nikolov_ephephrin_2013}. It is also known that that some receptor sub-types may compete for ligand attachment \citep{fiore_regulation_2019}.
Additionally, EphA4 receptors make `side attachments' to clusters of EphA receptors, possibly regulating the cluster size \citep{nikolov_ephephrin_2013}. 
%
We focused on the idea that signal strength is related to receptor cluster size and defined $c_i$, the effective cluster size for receptor $i$. 
We set $c_i=1$ for $i=1$, $2$ or $3$ and investigated a simple relationship between EphA cluster size, $c_0$, and $r_{\!\scriptscriptstyle A4}$, the density of EphA4 receptors:
\begin{equation}
    c_0 = 1/r_{\!\scriptscriptstyle A4}
\end{equation}
%
Eq.\,\ref{e:G} was modified to include the cluster size:
%
\begin{equation}\label{e:Gcs}
\mathbf{G} = m_{\!_G}\,\sum_i^N F_i\,c_i\,r_{i,b} \nabla L_i(\mathbf{x})
\end{equation}

@@

This study ignores the mechanism in species X, by which axons grow the full length of the tectum, then make branches at an appropriate location.
